Evolution of \\ ice sheet geometry \\ using Stokes dynamics

### 2021-06-12 talk/talk.tex
Few more edits.

\vspace{-1mm}
\begin{itemize}
\item glacier geometry which evolves according to climate
\item take simplest such problem (grounded, nonsliding, isothermal)
\item for simplicity, assume time-independent data on fixed $\Omega \subset \RR^2$:
    \begin{itemize}
    \item (net) climatic mass balance $a$ \hfill $\gets$ \emph{$a>0$ for accumulation}
    \item bed elevation $b$
    \end{itemize}
\item \emph{to compute}: surface elevation $s$ on $\Omega$ and icy domain $\Lambda_s \subset \RR^3$
\end{itemize}
\end{frame}

@@

\bigskip
\begin{columns}
    \column{0.55\textwidth}
        \begin{itemize}
        \item assumed for this talk:
            \begin{itemize}
            \item $s$ well-defined (no overhangs)
            \item $s=b$ where ice is not present
            \item $s$ is defined on all of $\Omega$
            \end{itemize}
        \end{itemize}
    \column{0.45\textwidth}
        \includegraphics[width=1.0\textwidth]{figs/simpledomain.png}
\end{columns}
\end{frame}

@@

    \begin{itemize}
    \item Glen-law effective viscosity with $\text{p}=(1/\text{n})+1(=4/3)$:
      $$\nu_\eps = \frac{1}{2} B_n \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$$
    \item \emph{count the 3 nonlinearities}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item almost no one is solving the implicit IGP
    \begin{itemize}
    \item except Bueler (2016), Brinkerhoff et al (2017) (\emph{shallow})
    \item and Wirbel \& Jarosch (2020) (\emph{semi-coupled, fake ice layer})
    \item none are scalable (\emph{single level})
    \end{itemize}
\item what are people doing instead?
    \begin{itemize}
    \item \alert{explicit time-stepping with $s \ge b$ enforced by truncation}
    \item often with shallow approximations, e.g.~PISM run below

@@

\begin{itemize}
\item to compute $\Phi(s)$: build $\Lambda_s$, solve weak-form Stokes problem
    $$F_{\Lambda_s}(\bu,p)[\bv,q] = \int_{\Lambda_s} 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0,$$
extract trace $\bus$, then extend $\Phi(s) = -\bus \cdot \bn_s$ by zero to $\Omega$
\item regarding Stokes problem for fixed geometry:
    \begin{itemize}
    \item well-posed over $W_0^{1,\pp}(\Lambda_s)^3 \times L^\qq(\Lambda_s)$ (Jouvet \& Rappaz 2011)
    \item optimal solver exists (Isaac et al 2015)
    \end{itemize}
\item $\therefore$ \, $\phi(s)$ is well-defined if $s$ is piecewise $C^1$
    \begin{itemize}
    \item a regularity conjecture
    \end{itemize}
\end{itemize}

@@

\begin{itemize}
\small
\item $\Phi : \mathcal{K} \subset W^{1,\qq}(\Omega) \to W^{-1,\pp}(\Omega)$ where $\qq=4$ and $\pp=4/3$
\item the Stokes stress balance is nonlocal so $\Phi$ is ``dense''
    \begin{itemize}
    \item $\Phi$ is some kind of integral operator? (\emph{convolve with Stokeslets})
    \item $\Phi$ is short range? (\emph{concentrated near the diagonal})
    \end{itemize}
\item $\Phi \approx \Phi_{\text{SIA}}$ for shallow glaciers, a differential operator:
    $$\Phi_{\text{SIA}}(s) = - \frac{\gamma}{\qq} (s-b)^{\qq} |\grad_2 s|^{\qq} - \grad_2 \cdot\left[\frac{\gamma}{\qq+1} (s-b)^{\qq+1} |\grad_2 s|^{\qq-2} \grad_2 s\right]$$

    \begin{itemize}
    \item $\Phi$ is diffusive in the large!
    \item $\Phi$ is degenerate at the ice margin!
    \end{itemize}
\item yet common to say $\Phi$ is hyperbolic because SKE is equivalent to a transport equation for thickness: $H_t + \overline{\bu}\, \grad H = a$
\end{itemize}
\end{frame}

>>>>

\vspace{-1mm}
\begin{itemize}
\item glacier geometry which evolves according to climate
    \begin{itemize}
    \item simplest such problem (grounded, nonsliding, isothermal)
    \end{itemize}
\item assume time-independent data on fixed $\Omega \subset \RR^2$:
    \begin{itemize}
    \item (net) climatic mass balance $a$ \hfill $\gets$ \emph{$a>0$ for accumulation}
    \item bed elevation $b$
    \end{itemize}
\item \emph{to compute}: surface elevation $s$ on $\Omega$ and icy domain $\Lambda_s \subset \RR^3$
\end{itemize}
\end{frame}

@@

\bigskip
\begin{columns}
    \column{0.55\textwidth}
        \begin{itemize}
        \item assumed for this talk:
            \begin{itemize}
            \item $s$ well-defined (no overhangs)
            \item $s=b$ where ice is not present

                \begin{itemize}
                \item[{\color{black} $\circ$}] $s$ is defined on all of $\Omega$
                \end{itemize}
            \end{itemize}
        \end{itemize}
    \column{0.45\textwidth}
        \includegraphics[width=1.0\textwidth]{figs/simpledomain.png}
\end{columns}
\end{frame}

@@

    \begin{itemize}
    \item Glen-law effective viscosity with $\text{p}=(1/\text{n})+1(=4/3)$:
      $$\nu_\eps = \frac{1}{2} B_n \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$$
    \item \alert{\emph{the 3 nonlinearities}}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item almost no one is solving the implicit IGP
    \begin{itemize}
    \item except Bueler (2016), Brinkerhoff et al (2017) (\emph{shallow})
    \item and Wirbel \& Jarosch (2020) (\emph{semi-coupled, fake ice layer})
    \item none are scalable (\emph{single level})
    \end{itemize}
\item what are people doing instead, to get science?
    \begin{itemize}
    \item \alert{explicit time-stepping with $s \ge b$ enforced by truncation}
    \item often with shallow approximations, e.g.~PISM run below

@@

\begin{itemize}
\item to compute $\Phi(s)$: build $\Lambda_s$, solve weak-form Stokes problem
    $$F_{\Lambda_s}(\bu,p)[\bv,q] = \int_{\Lambda_s} 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0,$$
extract trace $\bus$, then extend $\Phi(s) = -\bus \cdot \bn_s$ by zero to $\Omega$
\item regarding Stokes problem for fixed geometry:
    \begin{itemize}
    \item well-posed over $W_0^{1,\pp}(\Lambda_s)^3 \times L^\qq(\Lambda_s)$ (Jouvet \& Rappaz 2011)
    \item optimal solver exists (Isaac et al 2015)
    \end{itemize}
\item $\therefore$ \, $\phi(s)$ is well-defined if $s$ is piecewise $C^1$
    \begin{itemize}
    \item which is a regularity conjecture
    \end{itemize}
\end{itemize}

@@

\begin{itemize}
\small
\item $\Phi : \mathcal{K} \subset W^{1,\qq}(\Omega) \to W^{-1,\pp}(\Omega)$ where $\qq=4$ and $\pp=4/3$
\item the Stokes stress balance is nonlocal so $\Phi$ is ``dense''
    \begin{itemize}
    \item $\Phi$ is some kind of integral operator (\emph{convolve with Stokeslets})
    \item $\Phi$ is short range (\emph{concentrated near the diagonal})
    \end{itemize}
\item $\Phi \approx \Phi_{\text{SIA}}$ for shallow glaciers, a differential operator:
    $$\Phi_{\text{SIA}}(s) = - \frac{\gamma}{\qq} (s-b)^{\qq} |\grad_2 s|^{\qq} - \grad_2 \cdot\left[\frac{\gamma}{\qq+1} (s-b)^{\qq+1} |\grad_2 s|^{\qq-2} \grad_2 s\right]$$

    \begin{itemize}
    \item $\Phi$ is diffusive in the large
    \item $\Phi$ is degenerate at the ice margin
    \end{itemize}
\item yet common to say
    \begin{itemize}
    \item $\Phi$ is hyperbolic because SKE is a transport equation for thickness: $H_t + \overline{\bu}\, \grad H = a$
    \end{itemize}
\end{itemize}
\end{frame}
